\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-15 coqimpl.tex
Save.

Family B2 extends B.
  Extend FInductive b : Set 
    := uu : b.
  Extend Family neg_handler. 
    Field uu : self_B2.b 
      := self_B2.uu.
  EndFamily. 
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
  (* Abstraction for Inductive Type *)
Module Type b_3 (self__B: EmptySig_4).
Parameter (b : Set). Parameter (tt ff : b). End b_3.      
Module Type b_6 := b_3 EmptyMod.
  (* Recursor to Set for b *)
Module b_rec_12 (self__B: b_6).
Definition __recursor_type_b_rec :=
  forall P : self__B.b -> Set,
  P self__B.tt -> P self__B.ff 
  -> forall __i : self__B.b, P __i. End b_rec_12. 
  (* Field B.neg_handler.ff *)
Module ff_24 (self__B: b_6) 
             (self__neg_handler: tt_23 self__B).
Definition ff : self__B.b := self__B.tt. End ff_24.
  (* Intermediate Module solely for type checking *)
Module v_33_34 (self__B: neg_handler_26).
Include b_rec_12 self__B.
Parameter (recursor_for_type_checking : __recursor_type_b_rec).
Definition term_for_type_checking :=
  recursor_for_type_checking (fun _ : self__B.b => self__B.b.)
    self__B.neg_handler.tt
    self__B.neg_handler.ff.  
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Example Code for Inductive Type}\label{fig:plugin-example2}
\end{figure}

Taking \cref{fig:plugin-example2} as an example, when we define inductive type "W", the following "Field b" can only know "W" is a type (of "Set") and there are at least two constructors for it, just like what we do in module type "W_3". We cannot export the eliminator of "W" inside "W_3" because doing so "b" (and "b_19") will not be able to be inherited to the context where "W" is extended with a third constructor. Since the inhabited recursor is generally asserting there are only two constructors in "W".

But we do have to provide a way to construct a recursor, and do type-checking (or exhaustiveness checking). Thus we need to extract a elimination principle for future usage. Here we show one exemplar result when extracting the inductive principle.

To construct a recursor, we make each handler

The key insight here is that, to make each recursor handler inheritable, we need to seal an abstraction around the inductive type; but to construct a recursor (and carry out the exhaustiveness checking), we need to break this abstraction and see the concrete definition of the inductive type. Our meta-theory need to handler these two seemingly contradicting ideas simultaneously.

Since the recursor is constructed knowing a concrete inductive type, thus not inheritable, for all the children, we have to reconstruct and override with a new recursor (acting like an extra exhaustiveness checking). This is affordable since each recursor handler doesn't need to be rechecked.

@@

% explain ftheorem wrapping this complicated recursor
We also provide the tool of `FTheorem' to wrap the proof-irrelevant recursor (i.e. when reasoning stuff in "Prop"). This tool can avoid most boiler-plate code when writing a standard recursor, and is also open to extension like `FRecursor'.

\textbf{Propositional Partial Recursor.}
Propositional Partial Recursor can prove injection and discrimination of the constructors.
Propositional Partial Recursor is good enough because it and its computational axiom can prove that the vanilla inductive type can "embed" into the extensible inductive type. 
(i.e. there will be an left inversion of that injection, which is also witnessing the fact that those "constructors" can really act as constructors) 
Thus 1. every future extension can support this partial recursion; 2. every type support this partial recursor with its computational axiom at least "support these constructors" (because of the embedding).

>>>>

Family B2 extends B.
  Extend FInductive b : Set 
    := uu : b.
  Extend Family neg_handler. 
    Field uu : self_B2.b 
      := self_B2.uu.
  EndFamily. 
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
  (* Abstraction for Inductive Type *)
Module Type b_3 (self_B: EmptySig_4).
Parameter (b : Set). Parameter (tt ff : b). End b_3.      
Module Type b_6 := b_3 EmptyMod.
  (* Recursor to Set for b *)
Module b_rec_12 (self_B: b_6).
Definition __recursor_type_b_rec :=
  forall P : self_B.b -> Set,
  P self_B.tt -> P self_B.ff 
  -> forall __i : self_B.b, P __i. End b_rec_12. 
  (* Field B.neg_handler.ff *)
Module ff_24 (self_B: b_6) 
             (self__neg_handler: tt_23 self_B).
Definition ff : self_B.b := self_B.tt. End ff_24.
  (* Intermediate Module solely for type checking *)
Module v_33_34 (self_B: neg_handler_26).
Include b_rec_12 self_B.
Parameter (recursor_for_type_checking : __recursor_type_b_rec).
Definition term_for_type_checking :=
  recursor_for_type_checking (fun _ : self_B.b => self_B.b)
    self_B.neg_handler.tt
    self_B.neg_handler.ff.  
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Example Code for Inductive Type}\label{fig:plugin-example2}
\end{figure}

Let's look at \cref{fig:plugin-example2}, the example of boolean and negation. When we define inductive type "b", the following "Field neg_handler.ff" can only know "b" is a type (of "Set") and there are at least two constructors for it, just like what we declare in ``the abstracted interface'' module type "b_3". We cannot export the eliminator of "b" inside "b_3" because doing so "neg_handler.ff" (and "ff_24") will not be able to be inherited to the context where "b" is extended with a third constructor, since the existence of recursor alone will assert that there are only two constructors in "b", not ``extensible'' at all!

But we do have to provide a way to construct a recursor, and execute type checking (or exhaustiveness checking). Thus we need to extract a elimination principle for future usage. Here we show one exemplar extraction of the recursor in \mintinline{Coq}{Set} as module "b_rec_12".

To construct a recursor, we make for each constructor a handler and aggregate them inside a family. The whole family "neg_handler" can be considered as a large matching case. Doing so, family inheritance can function as matching-case reuse. Then we use "FRecursor" to specify the motive and inductive type. Here "FRecursor" will feed Coq
a module "v_33_34" for doing type-checking -- what it does is basically assert the existence of the recursor and apply to see if handlers can pass the type checking by Coq.

Since the recursor is constructed knowing a concrete inductive type, thus not inheritable, for all the children, we have to reconstruct and override with a new recursor (acting like an extra exhaustiveness checking). This is affordable since each recursor handler doesn't need to be rechecked -- look at "B2.neg_handler", there is only one new handler for "uu" and the handlers from the parent "B.neg_handler.tt,B.neg_handler.ff" are already type-checked and inherited(thanks to the fact that they only rely on ``the abstracted interface'' module "b_3"). Of course, our plugin will ``inherits'' the recursor "neg" but what the plugin does is doing exhaustiveness checking, again.

@@

% explain ftheorem wrapping this complicated recursor
We also provide a command "FTheorem" when the users want to use tactic programming instead. This tool can avoid most boiler-plate code (e.g. specify the handler family) when writing a recursor like above, invoke proof interaction mode, allow tactic usage, and is also open to extension like "FRecursor".

To summarize, the key insight here is that, to make each recursor handler inheritable, we need to seal an abstraction around the inductive type (e.g. module "b_3"); but to construct a recursor (and carry out the exhaustiveness checking), we need to break this abstraction and see the concrete definition of the inductive type (e.g. module "b_rec_12"). Our meta-theory need to handler these two seemingly contradicting ideas simultaneously.

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref*{fig:plugin-example2} is definitely not expressive enough. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt <> self_B.ff}.

Propositional Partial Recursor can prove injection and discrimination of the constructors.
Propositional Partial Recursor is good enough because it and its computational axiom can prove that the vanilla inductive type can "embed" into the extensible inductive type. 
(i.e. there will be an left inversion of that injection, which is also witnessing the fact that those "constructors" can really act as constructors) 
Thus 1. every future extension can support this partial recursion; 2. every type support this partial recursor with its computational axiom at least "support these constructors" (because of the embedding).
